---
title: "Expression on Xenium base panel features in 13384 tumors"
author: "heinin"
date: "2024-02-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

### Packages and environment variables

```{r}

suppressPackageStartupMessages({
  library(cli)
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(workflowr)
  library(googlesheets4)
  library(UpSetR)})

setwd("/home/hnatri/ILD_ASE_Xenium/")
set.seed(9999)
options(ggrepel.max.overlaps = Inf)

# Colors, themes, cell type markers, and plot functions
source("/home/hnatri/GBM_Xenium/code/colors_themes.R")
source("/home/hnatri/GBM_Xenium/code/plot_functions.R")
source("/home/hnatri/GBM_Xenium/code/utilities.R")

```

### Panel features

```{r}

# https://docs.google.com/spreadsheets/d/14SQGru_Wj1AzzKhRb-ckivscl55xrr5Jw1hyjQim6zs/edit?usp=sharing
gs4_deauth()
panel_sheets  <- gs4_get("https://docs.google.com/spreadsheets/d/14SQGru_Wj1AzzKhRb-ckivscl55xrr5Jw1hyjQim6zs/edit?usp=sharing")
sheet_names(panel_sheets)
breast <- read_sheet(panel_sheets, sheet = "Breast cancer base panel")
brain <- read_sheet(panel_sheets, sheet = "Brain base panel")
multitissue_cancer <- read_sheet(panel_sheets, sheet = "Multi-tissue/Cancer base panel")
gbm <- read_sheet(panel_sheets, sheet = "Berens lab GBM panel")
custom <- read_sheet(panel_sheets, sheet = "Additional genes")

panel_gene_list <- list("breast" = breast$Name,
                        "brain" = brain$Gene,
                        "multitissue_cancer" = multitissue_cancer$Gene,
                        "gbm" = gbm$Name,
                        "custom" = custom$Gene)

panel_gene_list <- panel_gene_list[order(sapply(panel_gene_list, length))]

# Cell type annotations for genes in the brain panel
brain$`Annotation (Xenium)` %>% table() %>%
  as.data.frame() %>% 
  dplyr::rename(celltype = ".") %>%
  ggplot(aes(x = reorder(celltype, -Freq), y = Freq)) +
    geom_bar(stat = "identity")+
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
    xlab("Cell type") +
    ylab("# features")

```

### Import data

```{r}

seurat_object <- readRDS("/tgen_labs/banovich/BCTCSF/Heini/tumor_integrated_UPN109pre_noUPN208_soupX_snn_metadata_no4_7.rds")
DefaultAssay(seurat_object)

# UMAP with cell type annotations
DimPlot(seurat_object,
        group.by = "celltype",
        cols = tumor_celltype_col,
        reduction = "integrated_sct_umap",
        label = T,
        label.box = T,
        label.size = 3,
        repel = T,
        raster = T,
        raster.dpi = c(1024, 1024),
        pt.size = 3) +
  ggtitle("") +
  theme_classic() +
  manuscript_theme + 
  NoLegend() +
  NoAxes() +
  coord_fixed(1)

# Pseudobulk expression (CPM) across cell types
exp_data <- AggregateExpression(seurat_object,
                                assay = "RNA",
                                group.by = "celltype",
                                normalization.method = "LogNormalize",
                                scale.factor = 1e6)
exp_data <- as.matrix(exp_data$RNA)

# Porportion of cells expression and (scaled) average expression
p <- DotPlot(object = seurat_object,
             features = unique(unlist(panel_gene_list)),
             group.by = "celltype")
  
# Extracting values
df <- p$data

# Removing NaN values
if(nrow(df[which(is.nan(df$avg.exp.scaled)),])>0){
  df <- df[-which(is.nan(df$avg.exp.scaled)),]
}

# Which panel features are present in our annotations?
lapply(panel_gene_list, function(xx){
  (length(intersect(rownames(exp_data), xx))/length(xx))*100
})

```

### Overlap of panel features

```{r}

upset(fromList(panel_gene_list), order.by = "freq")

```

### Panel feature expression across cell types

```{r}

# A list of genes meeting the expression threshold
exp_list <- lapply(unique(df$id), function(xx){
  
  df %>% filter(id == xx,
                pct.exp > 0.05,
                avg.exp > 1) %>%
    dplyr::select(features.plot) %>%
    unlist() %>% as.character()
})
names(exp_list) <- unique(df$id)

exp_list <- exp_list[order(sapply(exp_list, length))]

```

### Panel features expressed in scRNAseq with >5% of cells expressing, avg. exp >1

```{r, fig.width=10, fig.height=6, warning=F, message=F}

upset(fromList(c(panel_gene_list, exp_list)),
      keep.order = TRUE,
      order.by = "freq",
      nsets = length(c(panel_gene_list, exp_list)),
      mb.ratio = c(0.4, 0.6))

bar_cols <- c(rep("deeppink3", length(panel_gene_list)),
              rep("aquamarine3", length(exp_list)))
names(bar_cols) <- c(names(panel_gene_list), names(exp_list))

upset(fromList(c(panel_gene_list, exp_list)),
      keep.order = TRUE,
      sets = c(names(panel_gene_list), names(exp_list)),
      order.by = "freq",
      nsets = length(c(panel_gene_list, exp_list)),
      nintersects = 60,
      sets.bar.color = bar_cols,
      mb.ratio = c(0.4, 0.6))

```

### Same with >1% expressing, avg. exp >1

```{r}

exp_list <- lapply(unique(df$id), function(xx){
  
  df %>% filter(id == xx,
                pct.exp > 0.01,
                avg.exp > 1) %>%
    dplyr::select(features.plot) %>%
    unlist() %>% as.character()
})
names(exp_list) <- unique(df$id)

exp_list <- exp_list[order(sapply(exp_list, length))]

```


```{r, fig.width=10, fig.height=6, warning=F, message=F}

bar_cols <- c(rep("deeppink3", length(panel_gene_list)),
              rep("aquamarine3", length(exp_list)))
names(bar_cols) <- c(names(panel_gene_list), names(exp_list))

upset(fromList(c(panel_gene_list, exp_list)),
      keep.order = TRUE,
      sets = c(names(panel_gene_list), names(exp_list)),
      order.by = "freq",
      nsets = length(c(panel_gene_list, exp_list)),
      nintersects = 50,
      sets.bar.color = bar_cols,
      mb.ratio = c(0.4, 0.6))

```

### Looking at genes in intersects

#### Genes in the Barens lab GBM panel and expressed in any cell type

```{r}
# Intersect
intersect(panel_gene_list[["gbm"]], unlist(exp_list))

# Panel genes not expressed in scRNAseq
setdiff(panel_gene_list[["gbm"]], unlist(exp_list))

#gbm %>% filter(Name %in% setdiff(panel_gene_list[["gbm"]], unlist(exp_list)))

# % out of the whole panel expressed in scRNAseq
length(intersect(panel_gene_list[["gbm"]], unique(unlist(exp_list))))/length(panel_gene_list[["gbm"]])*100

```

#### Genes in the multi-tissue/cancer panel and expressed in any cell type

```{r}
# Intersect
intersect(panel_gene_list[["multitissue_cancer"]], unlist(exp_list))

# Panel genes not expressed in scRNAseq
setdiff(panel_gene_list[["multitissue_cancer"]], unlist(exp_list))

# Annotations for genes not expressed in scRNAseq
multitissue_cancer %>% filter(!Gene %in% unlist(exp_list))
  
# % out of the whole panel expressed in scRNAseq
length(intersect(panel_gene_list[["multitissue_cancer"]], unique(unlist(exp_list))))/length(panel_gene_list[["multitissue_cancer"]])*100

```

#### Genes in the brain panel and expressed in any cell type

```{r}
# Intersect
intersect(panel_gene_list[["brain"]], unlist(exp_list))

# Panel genes not expressed in scRNAseq
setdiff(panel_gene_list[["brain"]], unlist(exp_list))

# Annotations for genes not expressed in scRNAseq
brain %>% filter(!Gene %in% unlist(exp_list)) %>%
  select(`Annotation (Xenium)`) %>%
  table() %>%
  sort() %>%
  as.data.frame()

# % out of the whole panel expressed in scRNAseq
length(intersect(panel_gene_list[["brain"]], unique(unlist(exp_list))))/length(panel_gene_list[["brain"]])*100

```
